---
title: "BOurdeley et al. bulk RNA seq"
output: html_document
---
```{r}
library(DESeq2)
library(tidyverse)
library(GEOquery)
library(stringr)
library(EnhancedVolcano)
library(biomaRt)
library(dplyr)
```
```{r}
gse<-getGEO(GEO='GSE151073', GSEMatrix=TRUE)
data<-read_csv("GSE151073_PG_Bulk_Blood_raw_counts.csv.gz") #https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151073
```
```{r}

gse<-gse[[1]]
metadata <-pData(phenoData(gse))

```
```{r}
metadata1 <-dplyr::select(metadata,c(1,20,48,49,50))
metadata2 <- metadata1 %>%
  filter(title %in% c("DC3_1","DC3_2","DC3_3","DC3_4","DC2_CD5pos_1","DC2_CD5pos_2","DC2_CD5pos_3","DC2_CD5pos_4","DC2_TLR_1","DC2_TLR_3","DC2_TLR_4","DC3_TLR_1","DC3_TLR_3","DC3_TLR_4"))
data2 <-dplyr::select(data,c(1,6,7,8,9,10,11,12,13,28,29,30,31,32,33))
metadata3<-metadata2 %>% 
  mutate(across(where(is.character), str_remove_all, pattern = fixed(" ")))
metadata4<- metadata3 %>%
  dplyr::rename("celltype"="cell type:ch1") %>%
  dplyr::rename("description"="description.1")%>%
  dplyr::rename("donor"="donor:ch1")%>%
  dplyr::rename("tissue"="tissue:ch1")
metadata5 <- metadata4
vec <- c(rep("DC2_ctrl",4),rep("DC3_ctrl",4),rep("DC2_TLR",3), rep("DC3_TLR",3))

vec2 <- c(rep("DC2",4),rep("DC3",4),rep("DC2",3), rep("DC3",3))

vec3 <- c(rep("Ctrl",8),rep("TLR",6))
          
metadata6 <- cbind(metadata5,cell_group=vec,celltype2=vec2,stimulation=vec3)
rownames(metadata6) <- metadata6$description
data3 <- data2 %>%
  rename("GeneID"="...1")
```
```{r}
foo2 <- gsub("\\|.*", "", data3$GeneID)
data4 <- data3%>%
  mutate(GeneID=foo2)
foo3 <-gsub("\\.[0-9]*$","",data4$GeneID)
data5 <- data4 %>%
  mutate(GeneID=foo3)
```

```{r}
countdata <- data5 %>%
  column_to_rownames("GeneID") %>%
  as.matrix()
```
```{r}
keep <- rowSums(countdata) > 5
countdata <- countdata[keep,]
```
```{r}
librarySizes <- colSums(countdata)
bplot<-barplot(librarySizes, 
        names=names(librarySizes), 
        las=2, 
        cex.names=0.5,
        main="Barplot of library sizes")


```
#Problem?
```{r}
logcounts <- log2(countdata + 1)
statusCol <- match(metadata6$cell_group, c("DC2_ctrl", "DC3_ctrl", "DC2_TLR", "DC3_TLR")) + 1
boxplot(logcounts, 
        xlab="", 
        ylab="Log2(Counts)",
        las=2,
        col=statusCol)
abline(h=median(as.matrix(logcounts)), col="blue")
```

```{r}
library(ggfortify)

rlogcounts <- rlog(countdata)
pcDat <- prcomp(t(rlogcounts))
autoplot(pcDat,
         data = metadata6, 
         colour="cell_group",
         size=4)
```
```{r}
design2 <- as.formula(~cell_group)
modelMatrix <- model.matrix(design2, data = metadata6)
modelMatrix
```


```{r}
ddsObj.raw2 <- DESeqDataSetFromMatrix(countData = countdata,
                                     colData = metadata6,
                                     design = design2)
```
```{r}
vstcounts2 <- vst(ddsObj.raw2, blind=TRUE)
plotPCA(vstcounts2, intgroup=c("cell_group"))
```


```{r}
ddsObj2 <- DESeq(ddsObj.raw2)
res <- results(ddsObj2, alpha=0.05)
head(res)
```
```{r}
plotDispEsts(ddsObj2)
```
```{r}
resultsNames(ddsObj2)
```
```{r}
rDC3vDC2_c<-results(ddsObj2,
                    name="cell_group_DC3_ctrl_vs_DC2_ctrl",
                    alpha=0.05)
rDC3vDC2_t<-results(ddsObj2,
                    contrast=c("cell_group","DC3_TLR","DC2_TLR"),
                    alpha=0.05)
summary(rDC3vDC2_c)
```

```{r}
summary(rDC3vDC2_t)
```
```{r}
DESeq2::plotMA(rDC3vDC2_c)
```
```{r}
DESeq2::plotMA(rDC3vDC2_t)
```

## For visualization, we gonna use fcshrink function of DESeq
```{r}
library(ashr)
ddsShrink_c <- lfcShrink(ddsObj2, coef="cell_group_DC3_ctrl_vs_DC2_ctrl",type="ashr")
ddsShrink_t <- lfcShrink(ddsObj2, contrast = c("cell_group","DC3_TLR","DC2_TLR"), res = rDC3vDC2_t,type="ashr")
```

```{r}
Shrink_c<-as.data.frame(ddsShrink_c) %>%
  rownames_to_column("GeneID")

Shrink_c <- Shrink_c %>%
  dplyr::rename(logFC=log2FoldChange, FDR=padj)


Shrink_t<-as.data.frame(ddsShrink_t)%>%
  rownames_to_column("GeneID")%>%
  dplyr::rename(logFC=log2FoldChange, FDR=padj)

summary(ddsShrink_c)
summary(ddsShrink_t)
```
##Okay for later gsea stuff we need the entrezID and so on, so We have to get geneIDs from biomaRt
```{r}
Shrink_c <- Shrink_c %>%
  mutate(ENS_ID=GeneID)
Shrink_t <- Shrink_t %>%
  mutate(ENS_ID=GeneID)

```
#lets get the IDs from biomaRt
```{r}
listEnsembl()
ensembl <- useEnsembl("genes")
datasets <-listDatasets(ensembl)
ensembl.con <- useMart("ensembl", dataset="hsapiens_gene_ensembl",
                       host="uswest.ensembl.org")
attr <- listAttributes(ensembl.con)
filterValues <- Shrink_c$ENS_ID
filterValues2 <- Shrink_c$GeneID
annot2 <- getBM(
  attributes=c('ensembl_gene_id', 'entrezgene_id', 'external_gene_name',"entrezgene_accession"),
  filters ="ensembl_gene_id",
  values = filterValues,
  mart=ensembl.con
  )
```
#annotate
```{r}
annot2 <- annot2 %>%
  dplyr::rename("ENS_ID"="ensembl_gene_id")
##Fix the one to many relationship issue
fixedDuplicates <- annot2 %>%  
    add_count(ENS_ID) %>%  
    filter(n>1) %>% 
    dplyr::select(-n) %>% 
    filter(entrezgene_accession==external_gene_name)
annot3 <- annot2 %>%  
    add_count(ENS_ID) %>%  
    filter(n==1) %>% 
    dplyr::select(-n) %>% 
    bind_rows(fixedDuplicates)
annotC <-Shrink_c%>% 
    left_join(annot3, by="ENS_ID")
annotT <- Shrink_t%>%
  left_join(annot3,by="ENS_ID")
```
```{r}
par(mfrow=c(1,2))
hist(annotC$pvalue)
hist(annotT$pvalue)
```
```{r}
par(mfrow=c(1,2))
ggplot(annotC, aes(x = log2(baseMean), y=logFC)) + 
    geom_point(aes(colour=FDR < 0.05), shape=20, size=0.5)
ggplot(annotT, aes(x = log2(baseMean), y=logFC)) + 
    geom_point(aes(colour=FDR < 0.05), shape=20, size=0.5)
```
##Volcanoplots
```{r Figure 2, fig.height=6, fig.width=8}
annotT3 <- annotT %>%
  na_if("")%>%
  na.omit%>%
  filter(baseMean>20)
annotT4<- annotT %>%
  na_if("")%>%
  na.omit

annotC3 <- annotC %>%
  na_if("")%>%
  na.omit%>%
  filter(baseMean>20)
annotC4<- annotC %>%
  na_if("")%>%
  na.omit
keyvals <- ifelse(
    annotT3$logFC < -1.5 & annotT3$FDR < 10e-2, 'red',
      ifelse(annotT3$logFC > 1.5 &annotT3$FDR < 10e-2, 'purple',
        'grey'))
  keyvals[is.na(keyvals)] <- 'grey'
  names(keyvals)[keyvals == 'red'] <- 'enriched in cDC2'
  names(keyvals)[keyvals == 'grey'] <- "mid"
  names(keyvals)[keyvals == 'purple'] <- 'enriched in DC3'
EnhancedVolcano(annotT3,
                lab = annotT3$external_gene_name,
                title = 'DC3 versus cDC2 after TLR stimulation',
                FCcutoff = 1.5,
                pCutoff = 10e-2,
                colCustom = keyvals,
                ylab="-log10(FDR)",
                selectLab=c("NLRP3","NLRP1","NLRC4","IL1B","IL18","GSDMD","CASP1","CASP4","CASP5","NLRP12", "MEFV","CD163","CD36","CASP8","NLRP6","PYCARD","FCGR1A","FCGR2B","CLEC10A"),
                boxedLabels = TRUE,
                labSize = 6.0,
                labCol = 'black',
                labFace = 'bold',
                drawConnectors = TRUE,
                colConnectors = 'black',
                widthConnectors = 0.5,
                x= 'logFC',
                y ="FDR",
                pointSize = 1
                )
```
##Smething is wrong! It should be the exact other way around. But the model matrices are correct. What diid go wrong? Did they mislabel their data?

```{r, Figure3, fig.height=6, fig.width=6}
keyvals2 <- ifelse(
    annotC3$logFC < -1.5 & annotC3$FDR < 10e-2, 'red',
      ifelse(annotC3$logFC > 1.5 &annotC3$FDR < 10e-2, 'purple',
        'grey'))
  keyvals2[is.na(keyvals2)] <- 'grey'
  names(keyvals2)[keyvals2 == 'red'] <- 'enriched in cDC2'
  names(keyvals2)[keyvals2 == 'grey'] <- "mid"
  names(keyvals2)[keyvals2 == 'purple'] <- 'enriched in DC3'
EnhancedVolcano(annotC3,
                lab = annotC3$external_gene_name,
                title = 'DC3 versus cDC2 control',
                FCcutoff = 1.5,
                pCutoff = 10e-2,
                x= 'logFC',
                boxedLabels = TRUE,
                colCustom = keyvals2,
                labSize = 6.0,
                labCol = 'black',
                labFace = 'bold',
                drawConnectors = TRUE,
                colConnectors = 'black',
                widthConnectors = 0.5,
                y ="FDR")
```
#Heatmap
```{r}
library(ComplexHeatmap)
library(circlize)

annotT3 <- annotT %>%
  na_if("")%>%
  na.omit%>%
  filter(baseMean>20)
annotT4<- annotT %>%
  na_if("")%>%
  na.omit

annotC3 <- annotC %>%
  na_if("")%>%
  na.omit%>%
  filter(baseMean>20)
annotC4<- annotC %>%
  na_if("")%>%
  na.omit

sigGenesT <- annotT %>% 
    top_n(300, wt=-FDR) %>% 
    pull("ENS_ID")
sigGenesC <- annotC %>% 
    top_n(300, wt=-FDR) %>% 
    pull("ENS_ID")

sigGenesT3 <- annotT3 %>%
  top_n(150, wt=-FDR)
sigGenesT3_pull<- sigGenesT3%>%
  pull("ENS_ID")
sigGenesT4 <- annotT4 %>%
  top_n(150, wt=-FDR)
sigGenesT4_pull<- sigGenesT4%>%
  pull("ENS_ID")

sigGenesC3 <- annotC3 %>%
  top_n(150, wt=-FDR)
sigGenesC3_pull<- sigGenesC3%>%
  pull("GeneID")
sigGenesC4 <- annotC4 %>%
  top_n(150, wt=-FDR)
sigGenesC4_pull<- sigGenesC4%>%
  pull("GeneID")

```





```{r}
plotDat_heat_T <- vst(ddsObj2)[sigGenesT3_pull,] %>% 
    assay()
rownames(plotDat_heat_T) <- sigGenesT3$external_gene_name

z.mat_T <- t(scale(t(plotDat_heat_T), center=TRUE, scale=TRUE))
plotDat_heat_C <- vst(ddsObj2)[sigGenesC3_pull,] %>% 
    assay()
rownames(plotDat_heat_C) <- sigGenesC3$external_gene_name
z.mat_C <- t(scale(t(plotDat_heat_C), center=TRUE, scale=TRUE))

```

```{r Figure4, fig.height=6, fig.width=5}
library(grid)
myPalette <- c("blue3", "ivory", "red3")
myRamp = colorRamp2(c(-2, 0, 2), myPalette)
hcDatT <- hclust(dist(z.mat_T))
cutGroups <- cutree(hcDatT, h=4)

ha1 = HeatmapAnnotation(df = colData(ddsObj2)[,c("cell_group")])
htT<-Heatmap(z.mat_T, name = "z-score",
        col = myRamp,            
        show_row_names = TRUE,
        cluster_columns = TRUE,
        cluster_rows = TRUE,
        row_names_gp = grid::gpar(fontsize = 2),
        split=cutGroups,
        show_column_names = TRUE,
        rect_gp = gpar(col = "darkgrey", lwd=0.5),
        top_annotation = ha1)
htT[,9:14]
#column_labels = c(rep("",8),"DC2_d4","DC2_d1","DC2_d3","DC3_d4","DC3_d3","DC3_d1")
```
```{r Figure5, fig.height=6, fig.width=5}
hcDatC <- hclust(dist(z.mat_C))
cutGroups <- cutree(hcDatC, h=4)

ha1 = HeatmapAnnotation(df = colData(ddsObj2)[,c("cell_group")])
htC<-Heatmap(z.mat_C, name = "z-score",
        col = myRamp,            
        show_row_names = TRUE,
        row_labels = sigGenesT3$external_gene_name,
        row_names_gp = grid::gpar(fontsize = 2),
        cluster_columns = TRUE,
        split=cutGroups,
        show_column_names = TRUE,
        rect_gp = gpar(col = "darkgrey", lwd=0.5),
        top_annotation = ha1)
htC[,1:8]
#column_labels = c("DC3_d4","DC3_d3","DC3_d2","DC3_d1","DC2_d1","DC2_d3","DC2_d4","DC2_d2",rep("",6)),
```
#Heatmap with specific genes
```{r Figure6, fig.height=3, fig.width=3}
spGenes<- c("ENSG00000125538","ENSG00000150782","ENSG00000104518","ENSG00000103490","ENSG00000091592","ENSG00000162711","ENSG00000091106","ENSG00000174885","ENSG00000142405","ENSG00000103313","ENSG00000137752","ENSG00000196954","ENSG00000137757","ENSG00000151414", "ENSG00000137265", "ENSG00000140968")
rowlabels <- c("IL1B","IL18","GSDMD","PYCARD","NLRP1","NLRP3","NLRC4","NLRP6","NLRP12","MEFV","CASP1","CASP4","CASP5","NEK7", "IRF4", "IRF8")
plotDat_heat_TS <- vst(ddsObj2)[spGenes,] %>% 
    assay()
rownames(plotDat_heat_TS) <- rowlabels
z.mat_TS <- t(scale(t(plotDat_heat_TS), center=TRUE, scale=TRUE))
hcDatTS <- hclust(dist(z.mat_TS))
cutGroupsS <- cutree(hcDatTS, h=4)

htTS<-Heatmap(z.mat_TS, name = "z-score",
        col = myRamp,
        row_labels = rowlabels,
        show_row_names = TRUE,
        cluster_columns = FALSE,
        split=cutGroupsS,
        rect_gp = gpar(col = "darkgrey", lwd=0.5),
        top_annotation = ha1)
htTS#[,9:14]
```


```{r}
hyperGenes<-c("ENSG00000125538","ENSG00000137752","ENSG00000104518","ENSG00000004139","ENSG00000131165","ENSG00000255112","ENSG00000130724","ENSG00000083937","ENSG00000115561", "ENSG00000254505","ENSG00000101421","ENSG00000086065","ENSG00000176108","ENSG00000147457")
rowlabels2 <- c("IL1B","CASP1","GSDMD","SARM","CHMP1A","CHMP1B","CHMP2A","CHMP2B","CHMP3","CHMP4A","CHMP4B","CHMP5","CHMP6","CHMP7")
plotDat_heat_T_hyper <- vst(ddsObj2)[hyperGenes,] %>% 
    assay()
rownames(plotDat_heat_T_hyper) <- rowlabels2
z.mat_T_hyper <- t(scale(t(plotDat_heat_T_hyper), center=TRUE, scale=TRUE))
hcDatT_hyper <- hclust(dist(z.mat_T_hyper))
cutGroups_hyper <- cutree(hcDatT_hyper, k=4)

htT_hyper<-Heatmap(z.mat_T_hyper, name = "z-score",
        col = myRamp,
        row_labels = rowlabels2,
        show_row_names = TRUE,
        cluster_columns = FALSE,
        split=cutGroups_hyper,
        rect_gp = gpar(col = "darkgrey", lwd=0.5),
        top_annotation = ha1)
htT_hyper[,9:14]
```
```{r}
htT_hyper[,1:14]
```
```{r}
markerGenes<-c("ENSG00000110448","ENSG00000135218","ENSG00000177575","ENSG00000170458","ENSG00000150337","ENSG00000158481","ENSG00000125538","ENSG00000115008","ENSG00000163220","ENSG00000163221","ENSG00000114013","ENSG00000019582","ENSG00000179583")
rowlabels3 <- c("CD5","CD36","CD163","CD14","FCGR1A","CD1C","IL1B","IL1A","S100A9","S100A12","CD86","CD74","CIITA")
plotDat_heat_marker <- vst(ddsObj2)[markerGenes,] %>% 
    assay()
rownames(plotDat_heat_marker) <- rowlabels3
z.mat_marker <- t(scale(t(plotDat_heat_marker), center=TRUE, scale=TRUE))
hcDat_marker <- hclust(dist(z.mat_marker))
cutGroups_marker <- cutree(hcDat_marker, h=4)

htT_marker<-Heatmap(z.mat_marker, name = "z-score",
        col = myRamp,
        row_labels = rowlabels3,
        show_row_names = TRUE,
        cluster_columns = FALSE,
        split=cutGroups_marker,
        rect_gp = gpar(col = "darkgrey", lwd=0.5),
        top_annotation = ha1)
htT_marker[,9:14]
```
```{r}
htT_marker[,1:14]
```

```{r}
install.packages("pheatmap")
library(pheatmap)
pheatmap(z.mat_T,main = "pheatmap default")


```




